---
title: "RNA-seq Analysis Tutorial: From Raw Reads to Biological Insight"
author: "Ipek Selcen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300)
```

# Introduction

This notebook provides a complete RNA-seq analysis workflow using data from **GEO accession GSE79018**. 

## Biological Context

This analysis compares gene expression between control and knockout conditions to identify differentially expressed genes and understand the biological processes affected by the treatment.

## Workflow Overview

1. **Quality Control**: Assess raw read quality
2. **Preprocessing**: Trim adapters and low-quality bases
3. **Alignment**: Map reads to reference genome
4. **Quantification**: Count reads per gene
5. **Differential Expression**: Identify DEGs using DESeq2
6. **Functional Analysis**: GO and KEGG enrichment

---

# Load Required Libraries

```{r libraries}
# Load packages
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ggrepel)
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)
library(DT)

# Set theme for plots
theme_set(theme_bw())
```

---

# Data Loading

## Load Count Matrix

```{r load-data}
# Load count data
counts <- read.table(
  "../results/counts/count_matrix_clean.txt",
  header = TRUE,
  row.names = 1,
  check.names = FALSE)

# Load metadata
metadata <- read.csv("../data/metadata.csv", stringsAsFactors = FALSE)
rownames(metadata) <- metadata$sample_id

# Display dimensions
cat(sprintf("Count matrix dimensions: %d genes Ă— %d samples\n", 
            nrow(counts), ncol(counts)))

# Display first few rows
head(counts, 5)
```

## Sample Metadata

```{r metadata}
# Display metadata
datatable(metadata, options = list(pageLength = 10))
```

---

# Quality Control

## Library Size Distribution

```{r library-size}
# Calculate library sizes
lib_sizes <- colSums(counts)

# Create data frame for plotting
lib_df <- data.frame(
  sample = names(lib_sizes),
  size = lib_sizes,
  condition = metadata$condition)

# Plot
ggplot(lib_df, aes(x = sample, y = size/1e6, fill = condition)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Sample",
    y = "Library Size (millions)",
    title = "Library Size Distribution") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
```

## Gene Detection

```{r gene-detection}
# Count genes with at least 1 read per sample
genes_detected <- colSums(counts > 0)

genes_df <- data.frame(
  sample = names(genes_detected),
  genes = genes_detected,
  condition = metadata$condition)

ggplot(genes_df, aes(x = sample, y = genes, fill = condition)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Sample",
    y = "Number of Genes Detected",
    title = "Gene Detection per Sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
```

---

# Differential Expression Analysis

## Create DESeq2 Object

```{r create-dds}
# Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = metadata,
  design = ~ condition)

# Pre-filtering
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

cat(sprintf("Genes remaining after filtering: %d\n", sum(keep)))

# Set reference level
dds$condition <- relevel(dds$condition, ref = "Control")
```

## Run DESeq2

```{r run-deseq2}
# Run differential expression analysis
dds <- DESeq(dds)

# Get results
res <- results(dds, contrast = c("condition", "Treatment", "Control"))

# Summary
summary(res, alpha = 0.05)
```

## Extract Results

```{r extract-results}
# Convert to data frame
res_df <- as.data.frame(res)
res_df$gene_id <- rownames(res_df)

# Add regulation classification
res_df$regulation <- "Not Significant"
res_df$regulation[res_df$log2FoldChange > 1 & res_df$padj < 0.05] <- "Up"
res_df$regulation[res_df$log2FoldChange < -1 & res_df$padj < 0.05] <- "Down"

# Count DEGs
deg_summary <- table(res_df$regulation)
print(deg_summary)

# Display top DEGs
top_degs <- res_df %>%
  filter(padj < 0.05 & abs(log2FoldChange) > 1) %>%
  arrange(padj) %>%
  head(20)

datatable(top_degs, options = list(pageLength = 10))
```

---

# Visualization

## PCA Plot

```{r pca}
# Variance stabilizing transformation
vsd <- vst(dds, blind = FALSE)

# PCA
pca_data <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 5) +
  geom_text_repel(aes(label = name), size = 3) +
  labs(
    x = paste0("PC1: ", percentVar[1], "% variance"),
    y = paste0("PC2: ", percentVar[2], "% variance"),
    title = "PCA Plot - Sample Clustering") +
  scale_color_brewer(palette = "Set1")
```

## Sample Correlation Heatmap

```{r sample-correlation}
# Calculate correlation
normalized_counts <- counts(dds, normalized = TRUE)
sample_cor <- cor(normalized_counts)

# Plot heatmap
pheatmap(
  sample_cor,
  color = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
  display_numbers = TRUE,
  number_format = "%.2f",
  main = "Sample Correlation Heatmap")
```

## MA Plot

```{r ma-plot}
plotMA(res, alpha = 0.05, main = "MA Plot", ylim = c(-5, 5))
```

## Volcano Plot

```{r volcano}
# Prepare data
volcano_data <- res_df
volcano_data$log10padj <- -log10(volcano_data$padj)

# Label top genes
top_genes <- volcano_data %>%
  filter(padj < 0.05 & abs(log2FoldChange) > 2) %>%
  arrange(padj) %>%
  head(15)

# Plot
ggplot(volcano_data, aes(x = log2FoldChange, y = log10padj)) +
  geom_point(aes(color = regulation), alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c(
    "Up" = "red",
    "Down" = "blue",
    "Not Significant" = "gray")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_text_repel(
    data = top_genes,
    aes(label = gene_id),
    size = 2.5,
    max.overlaps = 15) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value",
    title = "Volcano Plot - Treatment vs Control") +
  theme(legend.position = "top")
```

## Heatmap of Top DEGs

```{r heatmap-degs}
# Select top 50 DEGs
sig_genes <- res_df %>%
  filter(padj < 0.05 & abs(log2FoldChange) > 1) %>%
  arrange(padj)

top_50_genes <- head(sig_genes, 50)$gene_id

# Get normalized counts
heatmap_data <- normalized_counts[top_50_genes, ]

# Z-score transformation
heatmap_data_scaled <- t(scale(t(heatmap_data)))

# Annotation
annotation_col <- data.frame(
  Condition = metadata$condition,
  row.names = rownames(metadata)

# Plot
pheatmap(
  heatmap_data_scaled,
  annotation_col = annotation_col,
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  scale = "none",
  main = "Top 50 Differentially Expressed Genes",
  show_rownames = TRUE,
  fontsize_row = 6,
  cluster_cols = TRUE,
  cluster_rows = TRUE)
```

---

# Functional Enrichment Analysis

## Gene ID Conversion

```{r gene-conversion}
# Get significant genes
sig_gene_ids <- sig_genes$gene_id

# Remove version numbers
sig_gene_ids_clean <- gsub("\\..*", "", sig_gene_ids)

# Convert to Entrez IDs
entrez_ids <- mapIds(
  org.Mm.eg.db,
  keys = sig_gene_ids_clean,
  column = "ENTREZID",
  keytype = "ENSEMBL",
  multiVals = "first")

entrez_ids <- na.omit(entrez_ids)

cat(sprintf("Converted %d genes to Entrez IDs\n", length(entrez_ids)))
```

## GO Enrichment - Biological Process

```{r go-bp}
go_bp <- enrichGO(
  gene = entrez_ids,
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05,
  readable = TRUE)

if (!is.null(go_bp) && nrow(go_bp) > 0) {
  # Display top terms
  datatable(
    as.data.frame(go_bp)[1:20, ],
    options = list(pageLength = 10))
  
  # Dotplot
  dotplot(go_bp, showCategory = 20) +
    ggtitle("GO Biological Process Enrichment")
  
  # Barplot
  barplot(go_bp, showCategory = 20) +
    ggtitle("GO Biological Process Enrichment")
}
```

## KEGG Pathway Analysis

```{r kegg}
kegg <- enrichKEGG(
  gene = entrez_ids,
  organism = "mmu",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05)

if (!is.null(kegg) && nrow(kegg) > 0) {
  kegg_readable <- setReadable(kegg, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
  
  # Display results
  datatable(
    as.data.frame(kegg_readable)[1:20, ],
    options = list(pageLength = 10))
  
  # Dotplot
  dotplot(kegg, showCategory = 20) +
    ggtitle("KEGG Pathway Enrichment")
}
```

---

# Summary

## Key Findings

```{r summary}
cat("=== Analysis Summary ===\n\n")

cat(sprintf("Total genes analyzed: %d\n", nrow(res_df)))
cat(sprintf("Significant DEGs (FDR < 0.05, |log2FC| > 1): %d\n", 
            sum(res_df$regulation != "Not Significant")))
cat(sprintf("  - Upregulated: %d\n", sum(res_df$regulation == "Up")))
cat(sprintf("  - Downregulated: %d\n", sum(res_df$regulation == "Down")))

if (!is.null(go_bp) && nrow(go_bp) > 0) {
  cat(sprintf("\nEnriched GO BP terms: %d\n", nrow(go_bp)))
}

if (!is.null(kegg) && nrow(kegg) > 0) {
  cat(sprintf("Enriched KEGG pathways: %d\n", nrow(kegg)))
}
```

## Session Information

```{r session-info}
sessionInfo()
```

---

# Conclusions

This analysis identified differentially expressed genes between knockout and control conditions, revealing biological processes and pathways affected by the knockout. The results provide insights into the molecular mechanisms underlying the observed phenotype.


---

**Analysis Date:** `r Sys.Date()`  
**Author:** Ipek Selcen  
**GitHub:** [https://github.com/ipekselcen/rnaseq-tutorial](https://github.com/ipekselcen/rnaseq-tutorial)
